A Variant of G\"{o}del's Ontological Proof \\ in a Natural Deduction Calculus \\ (Draft)

### 2014-08-15 Papers/NaturalDeduction/GodProof-ND.tex

\usepackage{calculi}
\usepackage{theorems}
\usepackage{natbib}

@@

\author{Bruno Woltzenlogel Paleo, Annika Siders}

@@

The first contribution of this paper is the detailed formalization of Scott's version \cite{Scott} of G\"odel's ontological argument \cite{Goedel} (as shown in Section \ref{sec:scott}) in a natural deduction calculus (as defined in Section \ref{sec:calculus}). A natural deduction style \citep{gentzen,prawitz} was chosen mainly for three reasons. Firstly, presentations of G\"oedel's proof are typically either informal or formalized in other styles of calculi (e.g. Fitting's tableaux \cite{fitting} or Sobel's sentential modal calculus \cite{sobel2}). Therefore, a formalization in natural deduction is a valuable complement to the existing presentations. Secondly, natural deduction is a familiar style for many people, who can now benefit from a natural deduction presentation of G\"odel's ontological argument. Finally, as natural deduction is the style used by proof assistants such as Coq \cite{coq} and Isabelle \cite{isabelle}, the natural deduction formalizations have been easily checked step-by-step in Coq \cite{coqformalizations}.

@@

The natural deduction calculus used here has standard rules for propositional connectives and quantifiers, as shown in Figures \ref{fig:PropositionalRules} and \ref{fig:QuantifierRules}. 
To deal with classicality \ref{fig:Classical} and \ref{fig:NDK}. These rules

A \emph{derivation} is a directed acyclic graph whose nodes are formulas and whose edges correspond to applications of the inference rules shown in Figures . Parts of a derivation may be surrounded by boxes. A \emph{proof} is a derivation that additionally satisfies the following conditions:

beta reduction

@@

ToDo: check if disjunction rules are used in Goedel's proof. If not, we could define disjunction as we already do for negation and equivalence.

@@

\s\s
\infer[\vee_E]{C}{A \vee B & \infer*{C}{\infer{A}{}} & \infer*{C}{\infer{B}{}}}
\s\s
\infer[\vee_{I_1}]{A \vee B}{A}
\s\s
\infer[\vee_{I_2}]{A \vee B}{B}

@@

\s
\infer[\all_I]{\all x. A[x]}{ A[\alpha] }
\s
\infer[\all_E]{A[t]}{ \all x. A[x] }
\s\s
\infer[\ex_I]{\ex x. A[x]}{ A[t] }
\s
\infer[\ex_E]{A[\beta]}{ \ex x. A[x] }

@@

\textbf{eigen-variable conditions:}
if $\rho$ is a $\all_I$ inference eliminating a variable $\alpha$, then any occurrence of $\alpha$ in the proof should be an ancestor of the occurrence of $\alpha$ eliminated by $\rho$;
if $\rho$ is a $\ex_E$ inference introducing a variable $\beta$, then any occurrence of $\beta$ in the proof should be a descendant of the occurrence of $\beta$ introduced by $\rho$.

@@

\s\s\s\s
\infer[\pos_I]{\pos A}{w: \fbox{\infer*{A}{}} }
\s\s\s\s\s
\infer[\pos_E]{\omega: \fbox{ \infer*{}{A} } }{\pos A}

\begin{itemize}
\item 
%
\item \textbf{eigen-box conditions:} $\omega$ must be a fresh name for a box.
%
\item \textbf{boxed assumption condition:} any assumption should be discharged within the box where it is made.
%
\item \textbf{unboxed root condition:} the proof's root should not be inside any box.
\end{itemize}

\noindent
Double lines are used to abbreviate tedious propositional reasoning steps in the derivations. Dashed lines are used to refer to a proof shown elsewhere. Dotted lines are used to indicate folding and unfolding of definitions.

\subsection{Suitability for K}

We show that the proof system with boxed parts of derivations is equivalent to the system K. The modal system K consists of the axiom K and the necessitation rule N.

\begin{axiom}[The transitivity axiom K]
\label{K}
$$
\nec(A\imp B)\imp (\nec A\imp \nec B)
$$
\end{axiom}

\begin{axiom}[The necessitation rule, $\Box_I$]
\label{Necessitation}
If $A$ is a theorem, then $\nec A$ is a theorem.
\end{axiom}

\begin{lemma}
\label{systemK}
The axiom K is derivable in the system.
\end{lemma}

\begin{small}
\begin{prooftree}
\AXC{$\nec(A\imp B)^2 $}\RightLabel{$\nec_E$}
\UIC{$A\imp B$}
      \AXC{$\nec A ^1 $}\RightLabel{$\nec_E$}
      \UIC{$A$} \RightLabel{$\imp_E$}
   \BIC{$ B$} \RightLabel{$\nec_I$}
   \UIC{$ \nec B$} \RightLabel{$\imp_I^1$}
   \UIC{$\nec A\imp \nec B$} \RightLabel{$\imp_I^2$}
   \UIC{$\nec(A\imp B)\imp (\nec A\imp \nec B)$}
\end{prooftree}
\end{small}

\begin{lemma}
\label{systemK2}
Assuming the axiom K and the necessitation rule $\Box_I$, the open formula $\nec A $ and the existence of a derivation of $ B$ from the open assumption $ A$, then we can derive $\nec B$ without the rules for boxed parts of derivations.
\end{lemma}

\begin{small}
\begin{prooftree}
\AXC{$A^1 $}\noLine
\UIC{$\vdots$}\noLine
\UIC{$ B$} \RightLabel{$\imp_I^1$}
\UIC{$A\imp B$} \RightLabel{$\Box_I$}
\UIC{$\nec(A\imp B)$}
      \AXC{Axiom K}\dashedLine
      \UIC{$\nec(A\imp B)\imp (\nec A\imp \nec B)$} \RightLabel{$\imp_E$}
  \BIC{$\nec A\imp \nec B$}
        \AXC{$\nec A ^1 $}\RightLabel{$\imp_E$}
    \BIC{$\nec B$}
\end{prooftree}
\end{small}

>>>>

\usepackage{calculi}
\usepackage{theorems}
\usepackage[numbers]{natbib}

@@

\author{Annika Siders\inst{1} \and Bruno Woltzenlogel Paleo\inst{2}}

\authorrunning{A.\~Siders \and B.\~Woltzenlogel Paleo}

\institute{
  Helsinki, Finland\\
  \email{annika.siders@helsinki.fi}
  \and 
  Vienna, Austria \\
  \email{bruno@logic.at}
}

@@

The first contribution of this paper is a detailed formalization of Scott's version \cite{Scott} of G\"odel's ontological argument \cite{Goedel} (as shown in Section \ref{sec:scott}) in a natural deduction calculus (as defined in Section \ref{sec:calculus}). A natural deduction style \citep{gentzen,prawitz} was chosen mainly for three reasons. Firstly, presentations of G\"oedel's proof are typically either informal or formalized in other styles of calculi (e.g. Fitting's tableaux \cite{fitting} or Sobel's sentential modal calculus \cite{sobel2}). Therefore, a formalization in natural deduction is a valuable complement to the existing presentations. Secondly, it makes the ontological proof accessible to people who are familiarized with a natural deduction style. Finally, as natural deduction is the style used by proof assistants such as Coq \cite{coq} and Isabelle \cite{isabelle}, the natural deduction formalizations have been easily checked step-by-step in Coq \cite{coqformalizations}.

@@

The natural deduction calculus used here has standard rules for propositional connectives and quantifiers, as shown in Figures \ref{fig:PropositionalRules} and \ref{fig:QuantifierRules}. The extension to classical logic is achieved by adding a rule for double negation elimination, shown in Figure \ref{fig:Classical}. Finally, modal operators are handled by special rules that insert or remove formulas from boxes, as shown in Figure \ref{fig:NDK}. Apart from the use of labels and the dual rules for `$\pos$', these rules are essentially the modal rules from \cite{todo}. Beta-reduction is implicit; all rules are assumed modulo beta-reduction. A \emph{derivation} is a directed acyclic graph whose nodes are formulas and whose edges correspond to applications of the inference rules. A \emph{proof} of a formula $F$ is a derivation without open assumptions and having $F$ as root not inside any box.

Double lines are used to abbreviate tedious propositional reasoning steps in the derivations. Dashed lines are used to refer to an axiom or theorem with proof shown elsewhere. Dotted lines are used to indicate folding and unfolding of definitions. Furthermore, as it is inconvenient to draw boxes around large derivations in \LaTeX, formulas inside boxes are labeled with the names of the boxes surrounding them. Therefore, the boxes can be omitted without loss of information.

@@

% ToDo: check if disjunction rules are used in Goedel's proof. If not, we could define disjunction as we already do for negation and equivalence.

@@

% \vspace{2em}

% \s\s
% \infer[\vee_E]{C}{A \vee B & \infer*{C}{\infer{A}{}} & \infer*{C}{\infer{B}{}}}
% \s\s
% \infer[\vee_{I_1}]{A \vee B}{A}
% \s\s
% \infer[\vee_{I_2}]{A \vee B}{B}

@@

\s
\infer[\all_I]{\all x_{\tau}. A[x]}{ A[\alpha] }
\s
\infer[\all_E]{A[t]}{ \all x_{\tau}. A[x] }
\s\s
\infer[\ex_I]{\ex x_{\tau}. A[x]}{ A[t] }
\s
\infer[\ex_E]{A[\beta]}{ \ex x_{\tau}. A[x] }

@@

\begin{center}
\textbf{eigen-variable conditions:} \\
if $\rho$ is a $\all_I$ inference eliminating a variable $\alpha$, then any occurrence of $\alpha$ in the proof should be an ancestor of the occurrence of $\alpha$ eliminated by $\rho$; \\
if $\rho$ is a $\ex_E$ inference introducing a variable $\beta$, then any occurrence of $\beta$ in the proof should be a descendant of the occurrence of $\beta$ introduced by $\rho$.
\end{center}

@@

% \vspace{2em}

% \s\s\s\s
% \infer[\pos_I]{\pos A}{w: \fbox{\infer*{A}{}} }
% \s\s\s\s\s
% \infer[\pos_E]{\omega: \fbox{ \infer*{}{A} } }{\pos A}

\begin{center}
\textbf{eigen-box condition:}\\ 
$\omega$ must be a fresh name for a box. \\
\vspace{0.5em}
\textbf{boxed assumption condition:} \\
assumptions should be discharged within the box where they are made.
\end{center}

%\clearpage

\noindent
The calculus having only the rules shown in Figures \ref{fig:PropositionalRules}, \ref{fig:Classical} and \ref{fig:QuantifierRules} is named \ND. The calculus with the additional rules shown in Figure \ref{fig:NDK} is called \NDK.

%\clearpage

\subsection{Suitability for Rigid Higher-Order Modal Logic K}

Adding the modal rules results in a calculus that is suitable for the basic normal modal logic K.
In other words, {\NDK} is sound and complete relative to {\ND} extended with axiom K ($\nec(A\imp B)\imp (\nec A\imp \nec B)$) and the necessitation rule (which establishes that $\nec A$ is a theorem if $A$ is a theorem).

The straightforward combinations of the quantifier rules of {\ND} either with the modal rules of {\NDK} or with axiom K and the necessitation rule are suitable for a higher-order modal logic where constants and variables are \emph{rigid}. From the point of view of a \emph{possible worlds} semantics, rigidity means that their interpretation is independent of the world in which they are being interpreted. Rigidity is silently assumed by most works investigating the ontological argument, and is explicitly assumed here. Nevertheless, it should be noted that its adequacy has already been contested \cite{fitting}.

\begin{theorem}
\label{theorem:completeness}
{\NDK} is complete, relative to {\ND} extended with axiom K and the necessitation rule.
\end{theorem}
\begin{proof}
The necessitation rule can be immediately simulated with the $\nec_I$ rule. Axiom K can be derived in {\NDK} as shown below:

\begin{small}
\begin{prooftree}
\AXC{$\nec(A\imp B)^2 $}\RightLabel{$\nec_E$}
\UIC{$\omega: A\imp B$}
      \AXC{$\nec A ^1 $}\RightLabel{$\nec_E$}
      \UIC{$\omega: A$} \RightLabel{$\imp_E$}
   \BIC{$ \omega: B$} \RightLabel{$\nec_I$}
   \UIC{$ \nec B$} \RightLabel{$\imp_I^1$}
   \UIC{$\nec A\imp \nec B$} \RightLabel{$\imp_I^2$}
   \UIC{$\nec(A\imp B)\imp (\nec A\imp \nec B)$}
\end{prooftree}
\end{small}

\end{proof}

\begin{theorem}
\label{theorem:soundness}
{\NDK} is sound, relative to {\ND} extended with axiom K and the necessitation rule.
\end{theorem}
\begin{proof}
It is necessary to show that {\NDK} proofs of the following form can be translated to proofs in {\ND} extended with the axiom K and the necessitation rule.

\begin{small}
$$
\infer[\nec_I]{\nec B}{
\infer{\omega: B}{
  \infer{\vdots}{\infer[\nec_E]{\omega: A_1}{\nec A_1}} &
  \ldots & 
  \infer{\vdots}{\infer[\nec_E]{\omega: A_n}{\nec A_n}}}
}
$$
\end{small}

\noindent
A translation to {\ND} extended with axiom K and necessitation is shown below for the case when $n=1$:

 % Assuming the axiom K and the necessitation rule $\Box_I$, the open formula $\nec A $ and the existence of a derivation of $ B$ from the open assumption $ A$, then we can derive $\nec B$ without the rules for boxed parts of derivations.

\begin{small}
\begin{prooftree}
\AXC{$ $}\RightLabel{1}
\UIC{$A_1$}\noLine
\UIC{$\vdots$}\noLine
\UIC{$ B$} \RightLabel{$\imp_I^1$}
\UIC{$A_1\imp B$} \RightLabel{necessitation}
\UIC{$\nec(A_1\imp B)$}
      \AXC{Axiom K}\dashedLine
      \UIC{$\nec(A_1\imp B)\imp (\nec A_1\imp \nec B)$} \RightLabel{$\imp_E$}
  \BIC{$\nec A_1\imp \nec B$}
        \AXC{$\nec A_1 $}\RightLabel{$\imp_E$}
    \BIC{$\nec B$}
\end{prooftree}
\end{small}

\noindent
For $n > 1$, the translation is a straightforward generalization:

\begin{scriptsize}
\begin{prooftree}
\AXC{$ $}\RightLabel{1}
\UIC{$A_1$}\noLine
\UIC{$\vdots$}
%
\AXC{$\ldots$}
%
\AXC{$ $}\RightLabel{n}
\UIC{$A_n$}\noLine
\UIC{$\vdots$} 
%
\TIC{$ B$} \doubleLine \RightLabel{$\imp_I^*$}
\UIC{$A_1\imp \ldots \imp A_n\imp B$} \RightLabel{nec.}
\UIC{$\nec(A_1\imp \ldots \imp A_n\imp B)$}
      \AXC{Axiom K, iterated}\doubleLine\dashedLine
      \UIC{$\nec(A_1\imp \ldots \imp A_n\imp B)\imp (\nec A_1\imp \ldots \imp \nec A_n\imp \nec B)$} \RightLabel{$\imp_E$}
  \BIC{$\nec A_1\imp \ldots \imp \nec A_n\imp \nec B$}
\end{prooftree}
\end{scriptsize}

\begin{scriptsize}
\begin{prooftree}
  \AXC{$ $} \dashedLine
  \UIC{$\nec A_1\imp \ldots \imp \nec A_n\imp \nec B$}
        \AXC{$\nec A_1 \quad \ldots \quad \nec A_n$} \doubleLine \RightLabel{$\imp_E$}
    \BIC{$\nec B$}
\end{prooftree}
\end{scriptsize}

\end{proof}
